OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-04 paper/paper.Rmd
Write description of algorithm

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference. _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Unstable OTU assignments make it
difficult to use _de novo_ clustering to compare taxonomic composition of
communities between studies or to use machine learning models trained with _de
novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
approach, first performing closed reference clustering, then any sequences that
cannot be assigned to reference OTUs are clustered _de novo_ to create
additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
algorithm created the highest quality OTU assignments of all clustering methods
based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

@@

### OptiFit algorithm

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.

- **TODO: brief description of OptiFit algorithm**
- **TODO: open & closed reference modes**
- **TODO: MCC score; only considers query sequences (when printref=f)**

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode.

@@

- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
- TODO: avoid slightly/much -- use numbers, e.g. X% better/worse.
- TODO: put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed reference mode, OTU quality was slightly worse
when fitting to Greengenes and Silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. In terms of run time,
closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
open reference OptiFit.

>>>>

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented by a similar sequence in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference. _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Unstable OTU assignments make it
difficult to use _de novo_ clustering to compare taxonomic composition of
communities between studies or to use machine learning models trained with _de
novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
approach, first performing closed reference clustering, then any sequences that
cannot be assigned to reference OTUs are clustered _de novo_ to create
additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
algorithm created the highest quality OTU assignments of all clustering methods
based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

@@

### The OptiFit algorithm

- **TODO: ask Sarah Westcott to check the accuracy of this description**

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering. OptiFit takes as input a list of
reference OTUs and their sequences, a list of query sequences to assign to the
reference OTUs, the sequence pairs that are within the distance threshold (e.g.
0.03), and the metric to assess clustering quality (default: MCC). Query
sequences are randomly seeded in reference OTUs, then for each sequence the
algorithm calculates the quality metric based on whether the sequences stays in
its current OTU or moves to each of the other OTUs. This process is repeated
until the quality metric stabilizes, changing by no more than 0.0001 by default,
or until a maximum number of iterations is reached (default: 100). In
closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: what determines whether a seq can't be assigned?**), and the
results only contain OTUs that exist in the original reference. In
open-reference mode, un-assigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence distance threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode.

@@

- **TODO:** separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
- **TODO:** avoid slightly/much -- use numbers, e.g. X% better/worse.
- **TODO:** put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed reference mode, OTU quality was slightly worse
when fitting to Greengenes and Silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. In terms of run time,
closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
open reference OptiFit.
